---
title: "Modulair PM2.5 Model Development"
author: "Mason Dow"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

```{r setup, warning=FALSE, include=FALSE, echo=FALSE}
library(tidyverse)
library(tidymodels)
library(readxl)
library(lubridate)
library(AirMonitor)
library(here)
library(purrr)
library(knitr)

source(here("functions", "analysis_functions.R"))

theme_set(theme_minimal())

# For tidymodels, seed by date for reproducibility

set.seed(20250314)
```

# Load Data 

```{r}
# Load transformed hourly and daily data from R environment (generated in EDA Preprocessing) 

lt_data <- lt_data
nc_data <- nc_data
rp_data <- rp_data

lt_daily <- lt_daily
nc_daily <- nc_daily
rp_daily <- rp_daily

```

# Transform Data

```{r}
# Recode columns (& merge datasets, ignoring merge for now)
rp_modeling_data <- rp_data %>%
 rename(fem_avg = rp_fem_avg,
        mod_pm25 = rp_modulair_pm25,
        fem_aqi = rp_fem_aqi,
        mod_aqi = rp_modulair_aqi,
        mod_rh = rp_modulair_rh,
        mod_tempc = rp_modulair_tempc)

# #nc_combined <- nc_data %>%
#   rename(fem_avg = nc_1020,
#          mod_pm25 = nc_modulair_pm25,
#          fem_aqi = nc_fem_aqi,
#          mod_aqi = nc_modulair_aqi,
#          mod_rh = nc_modulair_rh)
# 
# lt_combined <- lt_data %>%
#   rename(fem_avg = lt_thermo,
#          mod_pm25 = lt_modulair_pm25,
#          fem_aqi = lt_fem_aqi,
#          mod_aqi = lt_modulair_aqi,
#          mod_rh = lt_modulair_rh)

combined_modeling_data <- rp_modeling_data # only RP for now, other datasets not fit for model training/testing

```

# Model Development and Training via Tidymodels 

## Split Data into Training/Testing

```{r}
data_split <- initial_split(combined_modeling_data, prop = 0.8)
training_data <- training(data_split)
testing_data  <- testing(data_split)
```

## Recipes

```{r}
# Recipe using only modulair PM2.5 (base recipe)
base_linear_recipe <- recipe(fem_avg ~ mod_pm25, data = training_data)

# Recipe with additive RH term
linear_rh_recipe <- recipe(fem_avg ~ mod_pm25 + mod_rh, data = training_data)

# Recipe with modulair PM2.5, RH, and temperature
linear_temp_rh_recipe <- recipe(fem_avg ~ mod_pm25 + mod_rh + mod_tempc, data = training_data)

# Recipe with interaction between modulair PM2.5 and RH
interaction_recipe <- recipe(fem_avg ~ mod_pm25 + mod_rh, data = training_data) %>%
  step_interact(terms = ~ mod_pm25:mod_rh)

# Recipe with full interactions among modulair PM2.5, RH, and temperature
full_interaction_recipe <- recipe(fem_avg ~ mod_pm25 + mod_rh + mod_tempc, data = training_data) %>%
  step_interact(terms = ~ mod_pm25:mod_rh + mod_pm25:mod_tempc + mod_rh:mod_tempc)

# Recipe adding a second-degree (quadratic) term for modulair PM2.5
quad_recipe <- recipe(fem_avg ~ mod_pm25, data = training_data) %>%
  step_poly(mod_pm25, degree = 2, options = list(raw = TRUE))

```

## Model Specifications

```{r}
# Not necessary for more than single spec as all specifications are linear, but can keep in code for future use/customization for other potential non-linear pollutant modeling

# Base linear regression model (using only mod_pm25)
linear_spec <- linear_reg() %>% 
  set_engine("lm")

# Linear model with modulair PM2.5 and RH
linear_rh_spec <- linear_reg() %>% 
  set_engine("lm")

# Linear model with modulair PM2.5, RH, and temperature
linear_temp_rh_spec <- linear_reg() %>% 
  set_engine("lm")

# Model for interaction (will be paired with the interaction_recipe)
interaction_spec <- linear_reg() %>% 
  set_engine("lm")

# Quadratic model specification (to be used with quad_recipe)
quad_spec <- linear_reg() %>% 
  set_engine("lm")

```

## Workflows 

```{r}
# Workflow for base linear model
wf_linear <- workflow() %>%
  add_model(linear_spec) %>%
  add_recipe(base_linear_recipe)

# Workflow for "Linear + RH" model
wf_linear_rh <- workflow() %>%
  add_model(linear_rh_spec) %>%
  add_recipe(linear_rh_recipe)

# Workflow for "Linear + RH + Temperature" model
wf_linear_temp_rh <- workflow() %>%
  add_model(linear_temp_rh_spec) %>%
  add_recipe(linear_temp_rh_recipe)

# Workflow for interaction model (mod_pm25 and mod_rh interaction)
wf_interaction <- workflow() %>%
  add_model(interaction_spec) %>%
  add_recipe(interaction_recipe)

# Workflow for full interaction model 
wf_full_interaction <- workflow() %>%
  add_model(interaction_spec) %>%
  add_recipe(full_interaction_recipe)

# Workflow for quadratic model
wf_quad <- workflow() %>%
  add_model(quad_spec) %>%
  add_recipe(quad_recipe)
```

```{r}
# Define WF list:

wf_list <- list(
  "Base Linear"           = wf_linear,
  "Linear + RH"           = wf_linear_rh,
  "Linear + RH + Temp"    = wf_linear_temp_rh,
  "Interaction"           = wf_interaction,
  "Full Interaction"      = wf_full_interaction,
  "Quadratic"             = wf_quad
)
```

## Model Training

```{r}
# Model Fitting, Predictions and Metrics by Workflow 
# For Loop using fit_evaluate_wf function from function file 
# Loops over the workflows to fit models, predict on the test set, and collect metrics.

modeling_results <- map2(wf_list, names(wf_list), ~ fit_evaluate_wf(.x, .y, training_data, testing_data))

# Combine performance metrics from all models into a single table
all_modeling_metrics <- map_dfr(modeling_results, "metrics")
```

## Append Predictions to Testing Dataset

```{r}
# Append modeled predictions to the testing dataset
# Update the testing set with predictions from each model w/ new column for each model's prediction

# For each workflow, add a new column to the testing_data
testing_data_updated <- testing_data
for(model_name in names(wf_list)) {
  # Fit the workflow on the training data first
  fit_mod <- wf_list[[model_name]] %>% fit(data = training_data)
  # Generate predictions and add as a new column named "pred_<model_name>"
  col_name <- paste0("pred_", gsub(" ", "_", tolower(model_name)))
  testing_data_updated <- append_predictions(fit_mod, testing_data_updated, col_name)
}
```

## Add Corresponding Predicted AQI Categories to Testing Dataset

```{r}
# Build predicted AQI category columns for each prediction column
testing_data_updated <- testing_data_updated %>%
  mutate(across(all_of(prediction_cols), 
                ~ aqiCategories(.x, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
                .names = "{.col}_aqi"))
```

## Define Predictions Columns & Corresponding AQI Category Columns

```{r}
# Define the names of the prediction columns in the updated testing data
prediction_cols <- c("pred_base_linear", 
                     "pred_linear_+_rh", 
                     "pred_linear_+_rh_+_temp", 
                     "pred_interaction",
                     "pred_full_interaction",
                     "pred_quadratic")

# Define the corresponding predicted AQI column names
pred_aqi_cols <- paste0(prediction_cols, "_aqi")
```

# Model Evaluations

## Key Model Performance Metrics

```{r}
# Comparison table of performance metrics (rounded + formatted)

all_modeling_metrics %>%
  mutate(across(where(is.numeric), ~ round(.x, 2))) %>%
  kable(caption = "Key Accuracy Metrics by Model (Testing Data")
```

## Residuals Plots 

```{r}
# Generate residual plots for each prediction column
residual_plots <- map(prediction_cols, function(pred_col) {
  testing_data_updated %>%
    mutate(residual = fem_avg - .data[[pred_col]]) %>%
    ggplot(aes(x = .data[[pred_col]], y = residual)) +
      geom_point(alpha = 0.5) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
      labs(title = paste("Residual Plot for", pred_col),
           x = paste("Predicted FEM PM2.5 (", pred_col, ")"),
           y = "Residual (FEM - Predicted)") +
      theme_minimal()
})

# Name each plot for easier reference
names(residual_plots) <- prediction_cols

# Print each residual plot
walk(residual_plots, print)
```

## Calculate Stratified Metrics

```{r}
# Loop over each pair of prediction and predicted AQI columns and calculate metrics using your calculate_all_metrics function
metrics_by_model <- map2(prediction_cols, pred_aqi_cols, function(pred_col, pred_aqi_col) {
  calculate_all_metrics(
    data = testing_data_updated, 
    fem_col = "fem_avg", 
    mod_col = pred_col, 
    fem_aqi_col = "fem_aqi", 
    mod_aqi_col = pred_aqi_col
  )
})

# Extract overall metrics into a single tibble and add a column for the model name
overall_metrics <- map_dfr(metrics_by_model, "overall") %>%
  mutate(model = prediction_cols)

# Extract stratified (by FEM AQI category) metrics into a single tibble
stratified_metrics <- map2_dfr(metrics_by_model, prediction_cols, function(x, model_name) {
  x$stratified %>% mutate(model = model_name)
})

# Print the overall performance metrics table 
overall_metrics %>% 
  knitr::kable(caption = "Overall Performance Metrics by Model")

# Print the stratified performance metrics table 
stratified_metrics %>% 
  knitr::kable(caption = "Performance Metrics Stratified by AQI Category and Model")
```


# Final Model

## Selection Process (**YOU ARE HERE**)

## Model Fitting and Predictions

### Applying Linear Model to Entire RP Dataset

```{r}
# Fit the linear model (baseline) on the entire combined_modeling_data dataset.
final_fit_linear <- wf_linear %>% 
  fit(data = combined_modeling_data)

# Generate predictions on the entire dataset and recalculate the AQI categories using the corrected values.
lm_final_predictions <- predict(final_fit_linear, new_data = combined_modeling_data) %>% 
  bind_cols(combined_modeling_data) %>%
  mutate(corrected_mod_aqi = aqiCategories(.pred, pollutant = "PM2.5", NAAQS = "PM2.5_2024"))

# Evaluate performance using yardstick metrics (RMSE, R², and MAE).
final_performance_metrics <- lm_final_predictions %>%
  metrics(truth = fem_avg, estimate = .pred) %>%
  filter(.metric %in% c("rmse", "rsq", "mae"))

final_performance_metrics %>% 
  knitr::kable(caption = "Performance Metrics (Entire Dataset) for Linear Model")

# Calculate overall and stratified accuracy metrics using our custom function.
final_accuracy_metrics <- calculate_all_metrics(
  data = lm_predictions,
  fem_col = "fem_avg",
  mod_col = ".pred",
  fem_aqi_col = "fem_aqi",
  mod_aqi_col = "corrected_mod_aqi"
)

# Display the overall accuracy metrics.
final_accuracy_metrics$overall %>% 
  knitr::kable(caption = "Overall Accuracy Metrics on Entire Dataset")

# Display the stratified accuracy metrics by AQI category.
knitr::kable(final_accuracy_metrics$stratified, caption = "Stratified Accuracy Metrics by AQI Category - Entire Dataset")
```

## Accuracy

## Coefficients and Equation

```{r}
# Extract the underlying model object using extract_fit_parsnip()
lm_model <- extract_fit_parsnip(final_fit_linear)$fit

# Get the coefficients from the linear model
coefs <- coef(lm_model)
print(coefs)

# Create readable equation using the correct coefficient names
model_equation <- paste0("Corrected PM2.5 = ",
                         round(coefs["(Intercept)"], 2), " + ",
                         round(coefs["mod_pm25"], 2), " * Modulair PM2.5 + ",
                         round(coefs["mod_rh"], 2), " * Modulair rh")
cat("Linear Model Equation:\n", model_equation, "\n")
```

## Comparison to V1 PurpleAir U.S - wide correction equation 

### Assessing PurpleAir US-wide correction equation for Modulair data:

```{r}
# Apply the U.S. correction equation to the Modulair measurements
combined_modeling_data <- combined_modeling_data %>%
  mutate(
    pa_us_correction = 0.524 * mod_pm25 - 0.0862 * mod_rh + 5.75,
    pa_corrected_mod_aqi = aqiCategories(pa_us_correction, pollutant = "PM2.5", NAAQS = "PM2.5_2024")
  )

# Calculate accuracy metrics for the raw Modulair data
raw_metrics <- calculate_all_metrics(
  data = combined_modeling_data,
  fem_col = "fem_avg",
  mod_col = "mod_pm25",
  fem_aqi_col = "fem_aqi",
  mod_aqi_col = "mod_aqi"
)

# Calculate accuracy metrics for the data corrected using the U.S. equation
corrected_metrics <- calculate_all_metrics(
  data = combined_modeling_data,
  fem_col = "fem_avg",
  mod_col = "pa_us_correction",
  fem_aqi_col = "fem_aqi",
  mod_aqi_col = "pa_corrected_mod_aqi"
)

# Print overall accuracy metrics for raw and corrected data
raw_metrics$overall %>% 
  knitr::kable(caption = "Overall Accuracy Metrics: Raw Modulair Data")

corrected_metrics$overall %>% 
  knitr::kable(caption = "Overall Accuracy Metrics: U.S. Correction Applied")

# Optionally, compare stratified metrics by AQI category as well
knitr::kable(corrected_metrics$stratified, caption = "Stratified Accuracy Metrics by AQI Category (Corrected Data)")

```
