---
title: "Modulair PM2.5 Model Development"
author: "Mason Dow"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction


```{r setup, warning=FALSE, include=FALSE, echo=FALSE}
library(tidyverse)
library(tidymodels)
library(readxl)
library(lubridate)
library(AirMonitor)
library(here)

source(here("functions", "analysis_functions.R"))

theme_set(theme_minimal())

# For tidymodels, seed by date for reproducibility.

set.seed(20250314)
```

# Introduction

# Load Data 

```{r}
# Load data from data folder 
lt_data <- read_excel("../data/LT_data.xlsx")
nc_data <- read_excel("../data/NC_data.xlsx")
rp_data <- read_excel("../data/RP_data.xlsx")
```

# Transform Data

```{r}
# Add AQI categories using AirMonitor's aqiCategories() function.
rp_data <- rp_data %>%
  mutate(rp_modulair_aqi = aqiCategories(rp_modulair_pm25, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         rp_purpleair_aqi = aqiCategories(rp_purpleair_cf1, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         rp_fem_aqi = aqiCategories(rp_fem_avg, pollutant = "PM2.5", NAAQS = "PM2.5_2024"))

nc_data <- nc_data %>%
  mutate(nc_modulair_aqi = aqiCategories(nc_modulair_pm25, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         nc_purpleair_aqi = aqiCategories(nc_purpleair_cf1, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         nc_fem_aqi = aqiCategories(nc_1020, pollutant = "PM2.5", NAAQS = "PM2.5_2024"))

lt_data <- lt_data %>%
  mutate(lt_modulair_aqi = aqiCategories(lt_modulair_pm25, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         lt_purpleair_aqi = aqiCategories(lt_purpleair_cf1, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         lt_fem_aqi = aqiCategories(lt_thermo, pollutant = "PM2.5", NAAQS = "PM2.5_2024"))
```

```{r}
# Recode columns (& merge datasets, ignoring merge for now)
rp_data <- rp_data %>%
  rename(mod_pm25 = rp_modulair_pm25,
         fem_aqi = rp_fem_aqi,
         mod_aqi = rp_modulair_aqi,
         mod_rh = rp_modulair_rh)  

# #nc_combined <- nc_data %>%
#   rename(fem_avg = nc_1020,
#          mod_pm25 = nc_modulair_pm25,
#          fem_aqi = nc_fem_aqi,
#          mod_aqi = nc_modulair_aqi,
#          mod_rh = nc_modulair_rh)
# 
# lt_combined <- lt_data %>%
#   rename(fem_avg = lt_thermo,
#          mod_pm25 = lt_modulair_pm25,
#          fem_aqi = lt_fem_aqi,
#          mod_aqi = lt_modulair_aqi,
#          mod_rh = lt_modulair_rh)

combined_modeling_data <- rp_data # only RP for now

```

# Model Development and Testing via Tidymodels 

## Split Data into Training/Testing

```{r}
data_split <- initial_split(combined_modeling_data, prop = 0.8)
training_data <- training(data_split)
testing_data  <- testing(data_split)
```

## Recipes

```{r}
# Base recipe using mod and rh as predictors for fem_avg
base_linear_recipe <- recipe(fem_avg ~ mod_pm25, data = training_data)

linear_rh_recipe <- recipe(fem_avg ~ mod_pm25 + mod_rh, data = training_data)

# Quadratic recipe: add polynomial term for 'mod'
quad_recipe <- base_linear_recipe %>%
  step_poly(mod_pm25, degree = 2, options = list(raw = TRUE))

# Quadratic + RH interaction recipe: create interaction term between PM2.5 and Rh
quad_rh_recipe <- quad_recipe %>%
  step_interact(terms = ~ mod_pm25 * mod_rh - mod_pm25 - mod_rh)

# more later

```

## Model Specifications

```{r}
# Linear regression model (Model A baseline)
linear_spec <- linear_reg() %>% 
  set_engine("lm")

# For quadratic model (Model B), the recipe above includes a polynomial term.
quad_spec <- linear_reg() %>%
  set_engine("lm")

# We can use the same model specification for the more complex recipes.
# Later, we might explore non-linear models via nls() or gam(), but for now we start with linear regression.

```


## Workflows 

```{r}
# Workflow for Model "Linear" (baseline: linear model with only Modulair PM2.5)
wf_linear <- workflow() %>%
  add_model(linear_spec) %>%
  add_recipe(base_linear_recipe)

# Workflow for Model Linear RH (baseline: linear model with Modulair PM2.5 and Modulair RH)
wf_linear_rh <- workflow() %>%
  add_model(linear_spec) %>%
  add_recipe(linear_rh_recipe)

# Workflow for Model B (quadratic: using quad_recipe)
wf_quad <- workflow() %>%
  add_model(quad_spec) %>%
  add_recipe(quad_recipe)

# Workflow for Model D/E (quadratic + RH interaction)
wf_quad_rh <- workflow() %>%
  add_model(quad_spec) %>%
  add_recipe(quad_rh_recipe)
```


## Model Training

```{r}
# Fit each workflow on the training data.
fit_linear <- wf_linear %>% fit(data = training_data) # base linear model
fit_linear_rh <- wf_linear_rh %>% fit(data = training_data) # linear model w/ rh
fit_quad   <- wf_quad %>% fit(data = training_data)
#fit_quad_rh <- wf_quad_rh %>% fit(data = training_data)

```

##  Model Prediction Runs

```{r}
# Generate predictions on the testing set using the base linear model.
lm_predictions <- predict(fit_linear, new_data = testing_data) %>% 
  bind_cols(testing_data) %>%
  # Recalculate AQI categories using the predicted (corrected) PM2.5 values.
  mutate(lm_corrected_mod_aqi = aqiCategories(.pred, pollutant = "PM2.5", NAAQS = "PM2.5_2024"))

# Predictions using the linear + RH model.
lm_rh_predictions <- predict(fit_linear_rh, new_data = testing_data) %>% 
  bind_cols(testing_data) %>%
  # Recalculate AQI categories using the predicted (corrected) PM2.5 values.
  mutate(lm_rh_corrected_mod_aqi = aqiCategories(.pred, pollutant = "PM2.5", NAAQS = "PM2.5_2024"))


```


## Model Evaluation 

```{r}
# Base Linear Model

# Evaluate performance using yardstick metrics: RMSE, R², and MAE.
performance_metrics <- lm_predictions %>%
  metrics(truth = fem_avg, estimate = .pred) %>%
  filter(.metric %in% c("rmse", "rsq", "mae"))

performance_metrics %>% 
  knitr::kable(caption = "Performance Metrics (80/20 Split) for Linear Model")

# Calculate overall and stratified accuracy metrics using our custom function.
testing_accuracy <- calculate_all_metrics(
  data = lm_predictions,
  fem_col = "fem_avg",
  mod_col = ".pred",
  fem_aqi_col = "fem_aqi",
  mod_aqi_col = "lm_corrected_mod_aqi"
)

# Display overall accuracy metrics for the testing set.
testing_accuracy$overall %>% 
  knitr::kable(caption = "Overall LM Correction Accuracy Metrics on Testing Dataset")

# Display stratified accuracy metrics by AQI category for the testing set.
knitr::kable(testing_accuracy$stratified, caption = "Stratified LM Correction Accuracy Metrics by AQI Category - Testing Dataset")

```

### Applying Linear Model to Entire RP Dataset

```{r}
# Fit the linear model (baseline) on the entire combined_modeling_data dataset.
final_fit_linear <- wf_linear %>% 
  fit(data = combined_modeling_data)

# Generate predictions on the entire dataset and recalculate the AQI categories using the corrected values.
lm_predictions <- predict(final_fit_linear, new_data = combined_modeling_data) %>% 
  bind_cols(combined_modeling_data) %>%
  mutate(corrected_mod_aqi = aqiCategories(.pred, pollutant = "PM2.5", NAAQS = "PM2.5_2024"))

# Evaluate performance using yardstick metrics (RMSE, R², and MAE).
final_performance_metrics <- lm_predictions %>%
  metrics(truth = fem_avg, estimate = .pred) %>%
  filter(.metric %in% c("rmse", "rsq", "mae"))

final_performance_metrics %>% 
  knitr::kable(caption = "Performance Metrics (Entire Dataset) for Linear Model")

# Calculate overall and stratified accuracy metrics using our custom function.
final_accuracy_metrics <- calculate_all_metrics(
  data = lm_predictions,
  fem_col = "fem_avg",
  mod_col = ".pred",
  fem_aqi_col = "fem_aqi",
  mod_aqi_col = "corrected_mod_aqi"
)

# Display the overall accuracy metrics.
final_accuracy_metrics$overall %>% 
  knitr::kable(caption = "Overall Accuracy Metrics on Entire Dataset")

# Display the stratified accuracy metrics by AQI category.
knitr::kable(final_accuracy_metrics$stratified, caption = "Stratified Accuracy Metrics by AQI Category - Entire Dataset")

```


## Assessing PurpleAir US-wide correction equation for Modulair data:

```{r}
# Apply the U.S. correction equation to the Modulair measurements
combined_modeling_data <- combined_modeling_data %>%
  mutate(
    pa_us_correction = 0.524 * mod_pm25 - 0.0862 * mod_rh + 5.75,
    pa_corrected_mod_aqi = aqiCategories(pa_us_correction, pollutant = "PM2.5", NAAQS = "PM2.5_2024")
  )

# Calculate accuracy metrics for the raw Modulair data
raw_metrics <- calculate_all_metrics(
  data = combined_modeling_data,
  fem_col = "fem_avg",
  mod_col = "mod_pm25",
  fem_aqi_col = "fem_aqi",
  mod_aqi_col = "mod_aqi"
)

# Calculate accuracy metrics for the data corrected using the U.S. equation
corrected_metrics <- calculate_all_metrics(
  data = combined_modeling_data,
  fem_col = "fem_avg",
  mod_col = "pa_us_correction",
  fem_aqi_col = "fem_aqi",
  mod_aqi_col = "pa_corrected_mod_aqi"
)

# Print overall accuracy metrics for raw and corrected data
raw_metrics$overall %>% 
  knitr::kable(caption = "Overall Accuracy Metrics: Raw Modulair Data")

corrected_metrics$overall %>% 
  knitr::kable(caption = "Overall Accuracy Metrics: U.S. Correction Applied")

# Optionally, compare stratified metrics by AQI category as well
knitr::kable(corrected_metrics$stratified, caption = "Stratified Accuracy Metrics by AQI Category (Corrected Data)")

```

## Data Visulizations for Corrected Modulair PM2.5 Data

```{r}

# LM

plot_rp <- plot_fit(lm_predictions, mod_col = ".pred", fem_col = "fem_avg", site_label = "RP")
print(plot_rp)
```

```{r}

# PA US-Wide Correction Equation

plot_rp <- plot_fit(combined_modeling_data, mod_col = "pa_us_correction", fem_col = "fem_avg", site_label = "RP")
print(plot_rp)
```
```{r}
# Extract the underlying model object using extract_fit_parsnip()
lm_model <- extract_fit_parsnip(final_fit_linear)$fit

# Get the coefficients from the linear model
coefs <- coef(lm_model)
print(coefs)

# Create readable equation using the correct coefficient names.
model_equation <- paste0("Corrected PM2.5 = ",
                         round(coefs["(Intercept)"], 2), " + ",
                         round(coefs["mod_pm25"], 2), " * Modulair PM2.5 + ",
                         round(coefs["mod_rh"], 2), " * Modulair rh")
cat("Linear Model Equation:\n", model_equation, "\n")


```

