---
title: "Exploratory Analysis of MT DEQ FEM/Modulair PM2.5 Collocation Data"
author: "Mason Dow"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

This document performs an exploratory analysis of collocated FEM BAM and Modulair PM2.5 datasets from three MT DEQ sites (RP, NC, and LT). The objectives are to:

- Generate summary tables of descriptive statistics (min, max, mean, standard deviation, etc.) for each parameter for each site.
- Visualize PM2.5 trends 
- Compute AQI categories for PM2.5 using the AirMontitor package.
- Calculate a comprehensive set of accuracy metrics (including R², MAE, RMSE, normalized RMSE, regression slope, and intercept), overall and by AQI category for   each site.

# Setup

```{r setup, message=FALSE, warning=FALSE, echo=FALSE}

# Load necessary libraries
library(tidyverse)
library(readxl)
library(lubridate)
library(scales)
library(knitr)
library(AirMonitor)  # for aqiCategories()
library(here)

source(here("functions", "analysis_functions.R"))

theme_set(theme_minimal())

# Color scheme for monitor types
monitor_colors <- c("1020" = "yellow", 
                    "1022" = "red", 
                    "Thermo" = "tan", 
                    "FEM Average" = "green",
                    "Modulair" = "orange", 
                    "PurpleAir" = "purple")

# Define recoding maps for each site to convert raw column names to monitor labels
rp_recode <- c("rp_1020" = "1020",
               "rp_1022" = "1022",
               "rp_thermo" = "Thermo",
               "rp_fem_avg" = "FEM Average",
               "rp_modulair_pm25" = "Modulair",
               "rp_purpleair_cf1" = "PurpleAir")

nc_recode <- c("nc_1020" = "1020",
               "nc_modulair_pm25" = "Modulair",
               "nc_purpleair_cf1" = "PurpleAir")

lt_recode <- c("lt_thermo" = "Thermo",
               "lt_modulair_pm25" = "Modulair",
               "lt_purpleair_cf1" = "PurpleAir")
```


## Load Data

```{r}
# Load data from data folder 
lt_data <- read_excel("../data/LT_data.xlsx")
nc_data <- read_excel("../data/NC_data.xlsx")
rp_data <- read_excel("../data/RP_data.xlsx")
```

## Descriptive Statistics

```{r}

rp_stats <- descriptive_stats(rp_data)
nc_stats <- descriptive_stats(nc_data)
lt_stats <- descriptive_stats(lt_data)

kable(rp_stats, caption = "Descriptive Statistics - RP Site")
kable(nc_stats, caption = "Descriptive Statistics - NC Site")
kable(lt_stats, caption = "Descriptive Statistics - LT Site")

```

## Data Visualizations 

### Boxplots of PM2.5 Concentration Distributions

```{r}
rp_stacked <- stack_monitor_data(rp_data, "datetime", names(rp_recode), rp_recode)
ggplot(rp_stacked, aes(x = monitor, y = pm25, fill = monitor)) +
  geom_boxplot() +
  scale_fill_manual(values = monitor_colors) +
  labs(title = "PM2.5 Distribution at RP Site", x = "Monitor", y = "PM2.5 (µg/m³)") +
  theme(legend.position = "none")

nc_stacked <- stack_monitor_data(nc_data, "datetime", names(nc_recode), nc_recode)
ggplot(nc_stacked, aes(x = monitor, y = pm25, fill = monitor)) +
  geom_boxplot() +
  scale_fill_manual(values = monitor_colors) +
  labs(title = "PM2.5 Distribution at NC Site", x = "Monitor", y = "PM2.5 (µg/m³)") +
  theme(legend.position = "none")

lt_stacked <- stack_monitor_data(lt_data, "datetime", names(lt_recode), lt_recode)
ggplot(lt_stacked, aes(x = monitor, y = pm25, fill = monitor)) +
  geom_boxplot() +
  scale_fill_manual(values = monitor_colors) +
  labs(title = "PM2.5 Distribution at LT Site", x = "Monitor", y = "PM2.5 (µg/m³)") +
  theme(legend.position = "none")

```

### Time Series Line Graphs of PM2.5 Concentrations by Site

```{r}
# RP Site
ggplot(rp_data, aes(x = datetime)) +
  geom_line(aes(y = rp_1020, color = "1020")) +
  geom_line(aes(y = rp_1022, color = "1022")) +
  geom_line(aes(y = rp_thermo, color = "Thermo")) +
  geom_line(aes(y = rp_modulair_pm25, color = "Modulair")) +
  geom_line(aes(y = rp_purpleair_cf1, color = "PurpleAir")) +
  scale_color_manual(values = monitor_colors) +
  labs(title = "PM2.5 Time Series at RP Site", x = "Datetime", y = "PM2.5 (µg/m³)", color = "Monitor Type")

# NC Site
ggplot(nc_data, aes(x = datetime)) +
  geom_line(aes(y = nc_1020, color = "1020")) +
  geom_line(aes(y = nc_modulair_pm25, color = "Modulair")) +
  geom_line(aes(y = nc_purpleair_cf1, color = "PurpleAir")) +
  scale_color_manual(values = monitor_colors) +
  labs(title = "PM2.5 Time Series at NC Site", x = "Datetime", y = "PM2.5 (µg/m³)", color = "Monitor Type")

# LT Site
ggplot(lt_data, aes(x = datetime)) +
  geom_line(aes(y = lt_thermo, color = "Thermo")) +
  geom_line(aes(y = lt_modulair_pm25, color = "Modulair")) +
  geom_line(aes(y = lt_purpleair_cf1, color = "PurpleAir")) +
  scale_color_manual(values = monitor_colors) +
  labs(title = "PM2.5 Time Series at LT Site", x = "Datetime", y = "PM2.5 (µg/m³)", color = "Monitor Type")

```

### Scatterplot w/ Line of Best Fit & 1:1 by Site

```{r}

plot_rp <- plot_fit(rp_data, mod_col = "rp_modulair_pm25", fem_col = "rp_fem_avg", site_label = "RP")
print(plot_rp)

plot_nc <- plot_fit(nc_data, mod_col = "nc_modulair_pm25", fem_col = "nc_1020", site_label = "NC")
print(plot_nc)

plot_lt <- plot_fit(lt_data, mod_col = "lt_modulair_pm25", fem_col = "lt_thermo", site_label = "LT")
print(plot_lt)

```


## Build AQI Categories by Site 

```{r}
# Add AQI categories using AirMonitor's aqiCategories() function.
rp_data <- rp_data %>%
  mutate(rp_modulair_aqi = aqiCategories(rp_modulair_pm25, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         rp_purpleair_aqi = aqiCategories(rp_purpleair_cf1, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         rp_fem_aqi = aqiCategories(rp_fem_avg, pollutant = "PM2.5", NAAQS = "PM2.5_2024"))

nc_data <- nc_data %>%
  mutate(nc_modulair_aqi = aqiCategories(nc_modulair_pm25, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         nc_purpleair_aqi = aqiCategories(nc_purpleair_cf1, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         nc_fem_aqi = aqiCategories(nc_1020, pollutant = "PM2.5", NAAQS = "PM2.5_2024"))

lt_data <- lt_data %>%
  mutate(lt_modulair_aqi = aqiCategories(lt_modulair_pm25, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         lt_purpleair_aqi = aqiCategories(lt_purpleair_cf1, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         lt_fem_aqi = aqiCategories(lt_thermo, pollutant = "PM2.5", NAAQS = "PM2.5_2024"))
```
 
## Accuracy Metrics

### Metrics by Site & AQI Category
 
```{r}
# Apply accuracy metrics functions to all three datasets

datasets <- list(rp = rp_data, nc = nc_data, lt = lt_data)
results <- list()

col_info <- list(
  rp = list(fem = "rp_fem_avg", mod = "rp_modulair_pm25", fem_aqi = "rp_fem_aqi", mod_aqi = "rp_modulair_aqi"),
  nc = list(fem = "nc_1020", mod = "nc_modulair_pm25", fem_aqi = "nc_fem_aqi", mod_aqi = "nc_modulair_aqi"),
  lt = list(fem = "lt_thermo", mod = "lt_modulair_pm25", fem_aqi = "lt_fem_aqi", mod_aqi = "lt_modulair_aqi")
)

for (site in names(datasets)) {
  results[[site]] <- calculate_all_metrics(
    data = datasets[[site]],
    fem_col = col_info[[site]]$fem,
    mod_col = col_info[[site]]$mod,
    fem_aqi_col = col_info[[site]]$fem_aqi,
    mod_aqi_col = col_info[[site]]$mod_aqi
  )
}

overall_table <- bind_rows(lapply(names(results), function(site) {
  results[[site]]$overall %>% mutate(site = site)
}))
overall_table <- overall_table %>% select(site, everything())
kable(overall_table, caption = "Overall Accuracy Metrics by Site (FEM vs. Modulair)")

for (site in names(results)) {
  print(kable(results[[site]]$stratified, caption = paste("Accuracy Metrics by AQI Category -", site, "Site")))
}
```
 
### Create Combined Dataset 
 
```{r}
# Combine datasets 

rp_combined <- rp_data %>%
  rename(
    fem_avg = rp_fem_avg,
    mod = rp_modulair_pm25,
    fem_aqi = rp_fem_aqi,
    mod_aqi = rp_modulair_aqi
  )

nc_combined <- nc_data %>%
  rename(
    fem_avg = nc_1020,
    mod = nc_modulair_pm25,
    fem_aqi = nc_fem_aqi,
    mod_aqi = nc_modulair_aqi
  )

lt_combined <- lt_data %>%
  rename(
    fem_avg = lt_thermo,
    mod = lt_modulair_pm25,
    fem_aqi = lt_fem_aqi,
    mod_aqi = lt_modulair_aqi
  )

combined_data <- bind_rows(rp_combined, nc_combined, lt_combined)

```

### Calculate Overall Cccuracy Metrics 

```{r}
# apply accuracy metrics functions to combined dataset

combined_metrics <- calculate_all_metrics(
  data = combined_data,
  fem_col = "fem_avg",
  mod_col = "mod",
  fem_aqi_col = "fem_aqi",
  mod_aqi_col = "mod_aqi"
)

# Display the overall metrics for the combined dataset
print(combined_metrics$overall)

# Display the stratified accuracy metrics by AQI category for the combined dataset
kable(combined_metrics$stratified, caption = "Accuracy Metrics by AQI Category - Combined Data")
```

