---
title: "Exploratory Analysis of MT DEQ FEM/Modulair PM2.5 Collocation Data"
author: "Mason Dow"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

Accurate air quality measurements—especially of fine particulate matter (PM2.5)—are critical for public health, regulatory decision-making, and environmental management. Recent advancements in low-cost sensors have enabled broader spatial and temporal coverage of air quality monitoring. However, these sensors often exhibit biases relative to regulatory-grade instruments, such as those using Federal Equivalent Method (FEM) or Federal Reference Method (FRM) protocols. Assessing and correcting these biases is essential to ensure sensor data can reliably support decision-making and public communication.

This document presents an exploratory analysis of collocated hourly PM2.5 data collected at three Montana DEQ monitoring sites (Rossiter Park [RP], NCORE [NC], and Lewistown [LT]) from QuantAQ's **Modulair** sensors, each compared to co-located FEM BAM monitors as a ground truth reference.

The objectives of this analysis are to:

- Generate summary statistics (min, max, mean, standard deviation, etc.) for PM2.5 and meteorological parameters across all sites.
- Visualize PM2.5 trends and concentration distributions by site and sensor type.
- Assign hourly AQI categories using the 2024 EPA breakpoints via the **AirMonitor** R package.
- Evaluate the accuracy of raw Modulair and EPA-corrected PurpleAir PM2.5 measurements using a suite of performance metrics (R², MAE, RMSE, NRMSE, CV, slope, intercept, and AQI match frequency), calculated both overall and stratified by AQI category.
- Benchmark these results against performance targets from the U.S. EPA and manufacturer-reported accuracy metrics published by QuantAQ.

These findings inform our downstream modeling effort, where we aim to develop a correction model to improve the accuracy of raw Modulair sensor readings under Montana conditions.

# Setup

```{r setup, message=FALSE, warning=FALSE, echo=FALSE}

# Load necessary libraries
library(tidyverse)
library(readxl)
library(lubridate)
library(scales)
library(knitr)
library(AirMonitor)  
library(here)
library(GGally)
library(patchwork)

source(here("functions", "R_analysis_modeling_functions_dow_20250409.R"))

theme_set(theme_minimal())
```


# Load Data

```{r}
# Load data from data folder 
lt_data <- read_excel("../data/LT_data.xlsx")
nc_data <- read_excel("../data/NC_data.xlsx")
rp_data <- read_excel("../data/RP_data.xlsx")
```

# Data Transformations

## Calculate 24 hr. Averages

```{r}
# uses averaging_threshold and daily_averages functions from function file
# applies 75% (18 hr.) threshold to hourly values to caclulate daily averages 

# Apply the function to each dataset:
rp_daily <- daily_averages(rp_data)
nc_daily <- daily_averages(nc_data)
lt_daily <- daily_averages(lt_data)

head(rp_daily)
```


## Color Code and Column Name Mapping by Monitor

```{r}
# Color scheme for graphing by monitor types

monitor_colors <- c("1020" = "#56B4E9", 
                    "1022" = "#D55E00", 
                    "Thermo" = "#009E73", 
                    "FEM Average" = "red",
                    "Modulair" = "#E69F00", 
                    "PurpleAir" = "purple")

# Define recoding maps for each site to convert raw column names to monitor labels

rp_recode <- c("rp_1020" = "1020",
               "rp_1022" = "1022",
               "rp_thermo" = "Thermo",
               "rp_fem_avg" = "FEM Average",
               "rp_modulair_pm25" = "Modulair",
               "rp_purpleair_cf1" = "PurpleAir")

nc_recode <- c("nc_1020" = "1020",
               "nc_modulair_pm25" = "Modulair",
               "nc_purpleair_cf1" = "PurpleAir")

lt_recode <- c("lt_thermo" = "Thermo",
               "lt_modulair_pm25" = "Modulair",
               "lt_purpleair_cf1" = "PurpleAir")
```

## Define AQI Breakpoints 

```{r}
# Standard US daily PM2.5 AQI breakpoints (24-hr, updated 2024)
# Good: 0.0 - 9.0
# Moderate: 9.1 - 35.4
# Unhealthy for Sensitive Groups: 35.5 - 55.4 
# Unhealthy: 55.5 - 125.4
# Very Unhealthy: 125.5 - 225.4
# Hazardous: 225.5 +

aqi_breaks  <- c(9, 35.4, 55.4, 125.4, 225.4)
aqi_colors  <- c("green", "yellow", "orange", "red", "purple")
```

## Calculate Corresponding AQI Categories by site and PM2.5 parameter 

```{r}
# Hourly Data

# Add AQI categories using AirMonitor's aqiCategories() function.

# RP Site
rp_data <- rp_data %>%
  mutate(
    rp_modulair_aqi = aqiCategories(rp_modulair_pm25, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
    rp_purpleair_aqi = aqiCategories(rp_purpleair_cf1, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
    rp_1020_aqi = aqiCategories(rp_1020, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
    rp_1022_aqi = aqiCategories(rp_1022, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
    rp_thermo_aqi = aqiCategories(rp_thermo, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
    rp_fem_avg_aqi = aqiCategories(rp_fem_avg, pollutant = "PM2.5", NAAQS = "PM2.5_2024")
  )

# NC Site
nc_data <- nc_data %>%
  mutate(
    nc_modulair_aqi = aqiCategories(nc_modulair_pm25, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
    nc_purpleair_aqi = aqiCategories(nc_purpleair_cf1, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
    nc_1020_aqi = aqiCategories(nc_1020, pollutant = "PM2.5", NAAQS = "PM2.5_2024")
  )

# LT Site
lt_data <- lt_data %>%
  mutate(
    lt_modulair_aqi = aqiCategories(lt_modulair_pm25, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
    lt_purpleair_aqi = aqiCategories(lt_purpleair_cf1, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
    lt_thermo_aqi = aqiCategories(lt_thermo, pollutant = "PM2.5", NAAQS = "PM2.5_2024")
  )

```

```{r}
# RP Daily
rp_daily <- rp_daily %>%
  mutate(
    rp_modulair_aqi = aqiCategories(rp_modulair_pm25, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
    rp_purpleair_aqi = aqiCategories(rp_purpleair_cf1, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
    rp_fem_aqi = aqiCategories(rp_fem_avg, pollutant = "PM2.5", NAAQS = "PM2.5_2024")
  )

# NC Daily
nc_daily <- nc_daily %>%
  mutate(
    nc_modulair_aqi = aqiCategories(nc_modulair_pm25, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
    nc_purpleair_aqi = aqiCategories(nc_purpleair_cf1, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
    nc_fem_aqi = aqiCategories(nc_1020, pollutant = "PM2.5", NAAQS = "PM2.5_2024")
  )

# LT Daily
lt_daily <- lt_daily %>%
  mutate(
    lt_modulair_aqi = aqiCategories(lt_modulair_pm25, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
    lt_purpleair_aqi = aqiCategories(lt_purpleair_cf1, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
    lt_fem_aqi = aqiCategories(lt_thermo, pollutant = "PM2.5", NAAQS = "PM2.5_2024")
  )
```

## Create Combined Dataset with Consitent Column Names

```{r}
# For all datasets, rename site-specific columns to common names via *_combined tables

# RP Site
rp_combined <- rp_data %>%
  rename(
    fem_avg = rp_fem_avg,               # FEM avg PM2.5
    mod_pm25 = rp_modulair_pm25,         # Modulair PM2.5
    purpleair_pm25 = rp_purpleair_cf1,   # PurpleAir PM2.5
    mod_rh = rp_modulair_rh,             # Modulair RH
    mod_tempc = rp_modulair_tempc,       # Modulair Temperature
    fem_aqi = rp_fem_avg_aqi,            # FEM AQI
    mod_aqi = rp_modulair_aqi,           # Modulair AQI
    purpleair_aqi = rp_purpleair_aqi     # PurpleAir AQI
  )

# NC Site
nc_combined <- nc_data %>%
  rename(
    fem_avg = nc_1020,
    mod_pm25 = nc_modulair_pm25,
    purpleair_pm25 = nc_purpleair_cf1,
    fem_aqi = nc_1020_aqi,
    mod_aqi = nc_modulair_aqi,
    purpleair_aqi = nc_purpleair_aqi,
    mod_rh = nc_modulair_rh,
    mod_tempc = nc_modulair_tempc
  )

# LT Site
lt_combined <- lt_data %>%
  rename(
    fem_avg = lt_thermo,
    mod_pm25 = lt_modulair_pm25,
    purpleair_pm25 = lt_purpleair_cf1,
    fem_aqi = lt_thermo_aqi,
    mod_aqi = lt_modulair_aqi,
    purpleair_aqi = lt_purpleair_aqi,
    mod_rh = lt_modulair_rh,
    mod_tempc = lt_modulair_tempc
  )

# Combine sites
combined_data <- bind_rows(rp_combined, nc_combined, lt_combined)

# Select and order columns cleanly
combined_data <- combined_data %>%
  select(siteid, datetime, fem_avg, mod_pm25, purpleair_pm25, mod_rh, mod_tempc, fem_aqi, mod_aqi, purpleair_aqi)

# Preview combined data
head(combined_data)

```


# Calculate Descriptive Statistics

```{r}
# Summarize observations by site, AQI Category 

# uses summarize_counts function from function file 
# adjust fem_col, mod_col, aqi_col as appropriate for each dataset.

# For RP site:
rp_summary <- summarize_counts(
  data     = rp_data,
  fem_col  = "rp_fem_avg",       
  mod_col  = "rp_modulair_pm25",
  aqi_col  = "rp_fem_avg_aqi" # using FEM average for AQI col at RP
)

# For NC site 
nc_summary <- summarize_counts(
  data     = nc_data,
  fem_col  = "nc_1020",
  mod_col  = "nc_modulair_pm25",
  aqi_col  = "nc_1020_aqi"
)

# For LT site:
lt_summary <- summarize_counts(
  data     = lt_data,
  fem_col  = "lt_thermo",
  mod_col  = "lt_modulair_pm25",
  aqi_col  = "lt_thermo_aqi"
)

# Combine site-level summaries into one table, print w/ kable
combined_summaries <- bind_rows(rp_summary, nc_summary, lt_summary)

kable(
  combined_summaries,
  caption = "Summary of Observations and AQI Category Counts/Percentages by Site"
)

```


```{r}
# build out descriptive statistics tables by site
# uses descriptive_stats function from function file 

rp_stats <- descriptive_stats(rp_data)
nc_stats <- descriptive_stats(nc_data)
lt_stats <- descriptive_stats(lt_data)

kable(rp_stats, caption = "Descriptive Statistics - RP Site")
kable(nc_stats, caption = "Descriptive Statistics - NC Site")
kable(lt_stats, caption = "Descriptive Statistics - LT Site")
```

# Data Visualizations and Explorations 

## Boxplots of PM2.5 Concentration Distributions

```{r}
# create boxplots of PM2.5 distributions by site and monitor
# uses stack_monitor_data function from function file

rp_stacked <- stack_monitor_data(rp_data, "datetime", names(rp_recode), rp_recode) # create "stacked" monitor data to invert columns and rows for plotting
ggplot(rp_stacked, aes(x = monitor, y = pm25, fill = monitor)) +
  geom_boxplot() +
  scale_fill_manual(values = monitor_colors) +
  labs(title = "PM2.5 Distribution at RP Site", x = "Monitor", y = "PM2.5 (ug/m3)") +
  geom_hline( # Add annotated line at each AQI breakpoint
    yintercept = aqi_breaks,
    color = aqi_colors,
    linetype = "dashed"
  ) +
  annotate("text", x = Inf, y = 12,
           label = "Good",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 35.4,
           label = "Moderate",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 55.4,
           label = "USG",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 150.4,
           label = "Unhealthy",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 250.4,
           label = "V. Unhealthy",
           vjust = -0.5, hjust = 1.1) +
  theme_minimal() +
  theme(legend.position = "none")

nc_stacked <- stack_monitor_data(nc_data, "datetime", names(nc_recode), nc_recode)
ggplot(nc_stacked, aes(x = monitor, y = pm25, fill = monitor)) +
  geom_boxplot() +
  scale_fill_manual(values = monitor_colors) +
  labs(title = "PM2.5 Distribution at NC Site", x = "Monitor", y = "PM2.5 (ug/m3)") +
  geom_hline( 
    yintercept = aqi_breaks,
    color = aqi_colors,
    linetype = "dashed"
  ) +
  annotate("text", x = Inf, y = 12,
           label = "Good",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 35.4,
           label = "Moderate",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 55.4,
           label = "USG",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 150.4,
           label = "Unhealthy",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 250.4,
           label = "V. Unhealthy",
           vjust = -0.5, hjust = 1.1) +
  theme_minimal() +
  theme(legend.position = "none")

lt_stacked <- stack_monitor_data(lt_data, "datetime", names(lt_recode), lt_recode)
ggplot(lt_stacked, aes(x = monitor, y = pm25, fill = monitor)) +
  geom_boxplot() +
  scale_fill_manual(values = monitor_colors) +
  labs(title = "PM2.5 Distribution at LT Site", x = "Monitor", y = "PM2.5 (ug/m3)") +
 geom_hline( 
    yintercept = aqi_breaks,
    color = aqi_colors,
    linetype = "dashed"
  ) +
  annotate("text", x = Inf, y = 12,
           label = "Good",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 35.4,
           label = "Moderate",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 55.4,
           label = "USG",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 150.4,
           label = "Unhealthy",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 250.4,
           label = "V. Unhealthy",
           vjust = -0.5, hjust = 1.1) +
  theme_minimal() +
  theme(legend.position = "none")

```

## Time Series Line Graphs of 24 hour averaged PM2.5 Concentrations by Site

```{r}

final_date <- as.Date("2025-02-27")

# RP Site

ggplot(rp_daily, aes(x = date)) +
  geom_line(aes(y = rp_1020, color = "1020")) +
  geom_line(aes(y = rp_1022, color = "1022")) +
  geom_line(aes(y = rp_thermo, color = "Thermo")) +
  geom_line(aes(y = rp_modulair_pm25, color = "Modulair")) +
  geom_line(aes(y = rp_purpleair_cf1, color = "PurpleAir")) +
  # add AQI lines
  geom_hline(
    yintercept = aqi_breaks,
    color = aqi_colors,
    linetype = "dashed"
  ) +
  annotate("text", x = final_date + days(50), y = 12,     label = "Good",         vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(50), y = 35.4,   label = "Moderate",     vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(50), y = 55.4,   label = "USG",          vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(50), y = 150.4,  label = "Unhealthy",    vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(50), y = 250.4,  label = "V. Unhealthy", vjust = -0.5, hjust = 1.1) +
  scale_color_manual(values = monitor_colors) +
  labs(
    title = "PM2.5 Time Series at RP Site",
    x = "Date",
    y = "PM2.5 (µg/m³)",
    color = "Monitor Type"
  ) +
  theme_minimal()

# NC Site
ggplot(nc_daily, aes(x = date)) +
  geom_line(aes(y = nc_1020, color = "1020")) +
  geom_line(aes(y = nc_modulair_pm25, color = "Modulair")) +
  geom_line(aes(y = nc_purpleair_cf1, color = "PurpleAir")) +
  geom_hline(
    yintercept = aqi_breaks,
    color = aqi_colors,
    linetype = "dashed"
  ) +
  annotate("text", x = final_date + days(10), y = 12,     label = "Good",         vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(10), y = 35.4,   label = "Moderate",     vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(10), y = 55.4,   label = "USG",          vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(10), y = 150.4,  label = "Unhealthy",    vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(10), y = 250.4,  label = "V. Unhealthy", vjust = -0.5, hjust = 1.1) +
  scale_color_manual(values = monitor_colors) +
  labs(
    title = "PM2.5 Time Series at NC Site",
    x = "Date",
    y = "PM2.5 (µg/m³)",
    color = "Monitor Type"
  ) +
  theme_minimal()

# LT Site
ggplot(lt_daily, aes(x = date)) +
  geom_line(aes(y = lt_thermo, color = "Thermo")) +
  geom_line(aes(y = lt_modulair_pm25, color = "Modulair")) +
  geom_line(aes(y = lt_purpleair_cf1, color = "PurpleAir")) +
  geom_hline(
    yintercept = aqi_breaks,
    color = aqi_colors,
    linetype = "dashed"
  ) +
  annotate("text", x = final_date + days(10), y = 12,     label = "Good",         vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(10), y = 35.4,   label = "Moderate",     vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(10), y = 55.4,   label = "USG",          vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(10), y = 150.4,  label = "Unhealthy",    vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(10), y = 250.4,  label = "V. Unhealthy", vjust = -0.5, hjust = 1.1) +
  scale_color_manual(values = monitor_colors) +
  labs(
    title = "PM2.5 Time Series at LT Site",
    x = "Date",
    y = "PM2.5 (µg/m³)",
    color = "Monitor Type"
  ) +
  theme_minimal()

```

## Scatterplot w/ Line of Best Fit & 1:1 by Site

```{r}
plot_rp <- plot_fit(rp_data, mod_col = "rp_modulair_pm25", fem_col = "rp_fem_avg", site_label = "RP")
print(plot_rp)

plot_nc <- plot_fit(nc_data, mod_col = "nc_modulair_pm25", fem_col = "nc_1020", site_label = "NC")
print(plot_nc)

plot_lt <- plot_fit(lt_data, mod_col = "lt_modulair_pm25", fem_col = "lt_thermo", site_label = "LT")
print(plot_lt)

```
## Correlation Matrix by Modeling Variable 

```{r}
# Create correlation matrix with all variables to be potentially included in modeling using combined dataset

# - fem_avg: FEM PM2.5 measurement (true value)
# - mod_pm25: Modulair PM2.5 measurement (raw)
# - mod_rh: Modulair relative humidity
# - mod_tempc: Modulair ambient temperature (°C)

model_vars <- combined_data %>%
  select(fem_avg, mod_pm25, mod_rh, mod_tempc)

# Calculate and visualize the correlation matrix using ggpairs
ggpairs(model_vars,
        lower = list(continuous = wrap("smooth", alpha = 0.3, size = 0.1)),
        upper = list(continuous = wrap("cor", size = 4)),
        diag = list(continuous = "barDiag")) +
  labs(title = "Correlation Matrix of Candidate Modeling Variables")
```

### Matrix Interpretation:

- Moderate correlation between reference and raw sensor PM2.5 (r ∼ 0.77) suggests the sensor is “in the ballpark” but needs a bias correction.

- Minimal linear correlation between PM2.5 and RH/temperature does not rule out the possibility that adding an RH or temperature term can significantly improve the final model—particularly if the sensor’s error depends on meteorology.

- Strong collinearity between temperature and humidity (r ~ −0.66) means we should be mindful about including both in a single linear regression. In some cases, one meteorological term (often RH) captures most of the sensor’s bias.
 
# Measures of Sensor Accuracy
 
## Overall Accuracy Metrics 

```{r}
# ---------------------------------------------------------------
# Updated: Overall and Stratified Accuracy Metrics for Combined Dataset
# ---------------------------------------------------------------

# Helper function for clean labeling
add_labels <- function(metrics_tbl, sensor_label) {
  metrics_tbl %>%
    mutate(Sensor = sensor_label) %>%
    relocate(Sensor, .before = everything())
}

# ----------------
# Overall Accuracy Metrics
# ----------------

# Modulair vs FEM
combined_modulair_overall <- calculate_all_metrics(
  data = combined_data,
  fem_col = "fem_avg",
  sensor_col = "mod_pm25",
  fem_aqi_col = "fem_aqi",
  sensor_aqi_col = "mod_aqi"
)

# PurpleAir vs FEM
combined_purpleair_overall <- calculate_all_metrics(
  data = combined_data,
  fem_col = "fem_avg",
  sensor_col = "purpleair_pm25",
  fem_aqi_col = "fem_aqi",
  sensor_aqi_col = "purpleair_aqi"
)

# Combine overall results
combined_overall_table <- bind_rows(
  add_labels(combined_modulair_overall, "Modulair"),
  add_labels(combined_purpleair_overall, "PurpleAir")
)

# Print overall table
kable(combined_overall_table, caption = "Overall Accuracy Metrics - Combined Dataset (Modulair and PurpleAir)")

# ----------------
# Stratified Accuracy Metrics
# ----------------

# Modulair vs FEM stratified
combined_modulair_stratified <- calculate_stratified_metrics(
  data = combined_data,
  fem_col = "fem_avg",
  sensor_col = "mod_pm25",
  fem_aqi_col = "fem_aqi"
)

# PurpleAir vs FEM stratified
combined_purpleair_stratified <- calculate_stratified_metrics(
  data = combined_data,
  fem_col = "fem_avg",
  sensor_col = "purpleair_pm25",
  fem_aqi_col = "fem_aqi"
)

# Combine stratified results
combined_stratified_table <- bind_rows(
  add_labels(combined_modulair_stratified, "Modulair"),
  add_labels(combined_purpleair_stratified, "PurpleAir")
)

# Print stratified table
kable(combined_stratified_table, caption = "Stratified Accuracy Metrics by AQI Category - Combined Dataset (Modulair and PurpleAir)")


```
## Metrics by Site and FEM Monitor:

```{r}

# Overall Accuracy Metrics by Site (Modulair and PurpleAir)

run_comparison <- function(data, fem_col, sensor_col, fem_aqi_col = NULL, sensor_aqi_col = NULL, sensor_label) {
  metrics <- calculate_all_metrics(data, fem_col, sensor_col, fem_aqi_col, sensor_aqi_col)
  
  if (is.list(metrics) && "overall" %in% names(metrics)) {
    metrics <- metrics$overall
  }
  
  metrics <- metrics %>%
    mutate(
      FEM_Reference = fem_col,
      Sensor = sensor_label
    )
  
  return(metrics)
}

# Initialize empty lists to store results
rp_metrics <- list()
nc_metrics <- list()
lt_metrics <- list()

# ----------------
# RP Site (compare to 3 FEMs and FEM average)
# ----------------

# Modulair comparisons at RP
rp_metrics[["Modulair_vs_1020"]]  <- run_comparison(rp_data, "rp_1020", "rp_modulair_pm25", "rp_1020_aqi", "rp_modulair_aqi", "Modulair")
rp_metrics[["Modulair_vs_1022"]]  <- run_comparison(rp_data, "rp_1022", "rp_modulair_pm25", "rp_1022_aqi", "rp_modulair_aqi", "Modulair")
rp_metrics[["Modulair_vs_Thermo"]] <- run_comparison(rp_data, "rp_thermo", "rp_modulair_pm25", "rp_thermo_aqi", "rp_modulair_aqi", "Modulair")
rp_metrics[["Modulair_vs_FEM_Avg"]] <- run_comparison(rp_data, "rp_fem_avg", "rp_modulair_pm25", "rp_fem_avg_aqi", "rp_modulair_aqi", "Modulair")

# PurpleAir comparisons at RP
rp_metrics[["PurpleAir_vs_1020"]]  <- run_comparison(rp_data, "rp_1020", "rp_purpleair_cf1", "rp_1020_aqi", "rp_purpleair_aqi", "PurpleAir")
rp_metrics[["PurpleAir_vs_1022"]]  <- run_comparison(rp_data, "rp_1022", "rp_purpleair_cf1", "rp_1022_aqi", "rp_purpleair_aqi", "PurpleAir")
rp_metrics[["PurpleAir_vs_Thermo"]] <- run_comparison(rp_data, "rp_thermo", "rp_purpleair_cf1", "rp_thermo_aqi", "rp_purpleair_aqi", "PurpleAir")
rp_metrics[["PurpleAir_vs_FEM_Avg"]] <- run_comparison(rp_data, "rp_fem_avg", "rp_purpleair_cf1", "rp_fem_avg_aqi", "rp_purpleair_aqi", "PurpleAir")

# Combine RP results
rp_metrics_table <- bind_rows(rp_metrics) %>%
  select(Sensor, FEM_Reference, everything())

# Print RP table
kable(rp_metrics_table, caption = "Overall Accuracy Metrics - RP Site")

# ----------------
# NC Site
# ----------------

# Modulair vs NC 1020
nc_metrics[["Modulair_vs_1020"]] <- run_comparison(nc_data, "nc_1020", "nc_modulair_pm25", "nc_1020_aqi", "nc_modulair_aqi", "Modulair")
# PurpleAir vs NC 1020
nc_metrics[["PurpleAir_vs_1020"]] <- run_comparison(nc_data, "nc_1020", "nc_purpleair_cf1", "nc_1020_aqi", "nc_purpleair_aqi", "PurpleAir")

# Combine NC results
nc_metrics_table <- bind_rows(nc_metrics) %>%
  select(Sensor, FEM_Reference, everything())

# Print NC table
kable(nc_metrics_table, caption = "Overall Accuracy Metrics - NC Site")

# ----------------
# LT Site
# ----------------

# Modulair vs LT Thermo
lt_metrics[["Modulair_vs_Thermo"]] <- run_comparison(lt_data, "lt_thermo", "lt_modulair_pm25", "lt_thermo_aqi", "lt_modulair_aqi", "Modulair")
# PurpleAir vs LT Thermo
lt_metrics[["PurpleAir_vs_Thermo"]] <- run_comparison(lt_data, "lt_thermo", "lt_purpleair_cf1", "lt_thermo_aqi", "lt_purpleair_aqi", "PurpleAir")

# Combine LT results
lt_metrics_table <- bind_rows(lt_metrics) %>%
  select(Sensor, FEM_Reference, everything())

# Print LT table
kable(lt_metrics_table, caption = "Overall Accuracy Metrics - LT Site")

```

## Metrics by AQI Category

### By Site

```{r}
# ---------------------------------------------------------------
# Stratified Accuracy Metrics by Site (using FEM Average at RP only)
# ---------------------------------------------------------------

# Helper function to run stratified comparison easily
run_stratified_comparison <- function(data, fem_col, sensor_col, fem_aqi_col, sensor_label) {
  metrics <- calculate_stratified_metrics(data, fem_col, sensor_col, fem_aqi_col)
  
  metrics <- metrics %>%
    mutate(
      Sensor = sensor_label,
      FEM_Reference = fem_col
    ) %>%
    relocate(Sensor, FEM_Reference, AQI_Category)  
  
  return(metrics)
}

# Initialize empty lists
rp_stratified_metrics <- list()
nc_stratified_metrics <- list()
lt_stratified_metrics <- list()

# ----------------
# RP Site (ONLY compare against FEM Average)
# ----------------

# Modulair vs FEM Average at RP
rp_stratified_metrics[["Modulair_vs_FEM_Avg"]] <- run_stratified_comparison(rp_data, "rp_fem_avg", "rp_modulair_pm25", "rp_fem_avg_aqi", "Modulair")
# PurpleAir vs FEM Average at RP
rp_stratified_metrics[["PurpleAir_vs_FEM_Avg"]] <- run_stratified_comparison(rp_data, "rp_fem_avg", "rp_purpleair_cf1", "rp_fem_avg_aqi", "PurpleAir")

# Combine RP stratified results
rp_stratified_table <- bind_rows(rp_stratified_metrics)

# Print RP table
kable(rp_stratified_table, caption = "Stratified Accuracy Metrics - RP Site (FEM Average Only)")

# ----------------
# NC Site
# ----------------

# Modulair vs NC 1020
nc_stratified_metrics[["Modulair_vs_1020"]] <- run_stratified_comparison(nc_data, "nc_1020", "nc_modulair_pm25", "nc_1020_aqi", "Modulair")
# PurpleAir vs NC 1020
nc_stratified_metrics[["PurpleAir_vs_1020"]] <- run_stratified_comparison(nc_data, "nc_1020", "nc_purpleair_cf1", "nc_1020_aqi", "PurpleAir")

# Combine NC stratified results
nc_stratified_table <- bind_rows(nc_stratified_metrics)

# Print NC table
kable(nc_stratified_table, caption = "Stratified Accuracy Metrics - NC Site")

# ----------------
# LT Site
# ----------------

# Modulair vs LT Thermo
lt_stratified_metrics[["Modulair_vs_Thermo"]] <- run_stratified_comparison(lt_data, "lt_thermo", "lt_modulair_pm25", "lt_thermo_aqi", "Modulair")
# PurpleAir vs LT Thermo
lt_stratified_metrics[["PurpleAir_vs_Thermo"]] <- run_stratified_comparison(lt_data, "lt_thermo", "lt_purpleair_cf1", "lt_thermo_aqi", "PurpleAir")

# Combine LT stratified results
lt_stratified_table <- bind_rows(lt_stratified_metrics)

# Print LT table
kable(lt_stratified_table, caption = "Stratified Accuracy Metrics - LT Site")

```

# Preliminary Conclusions from EDA and Rationale for Model Selection

This section summarizes key findings from the exploratory data analysis (EDA), organized into three parts: (1) PM₂.₅ distribution characteristics by site and sensor type, (2) overall and stratified accuracy of the Modulair sensor compared to FEM monitors, and (3) implications for model development and selection.

## 1. PM₂.₅ Distributions by Site and Sensor

- **RP Site:**
  - Displays the widest range of PM₂.₅ values across all sensors given the longer collocation period, with the Modulair reaching ~170 µg/m³ and the FEM average peaking at ~105 µg/m³.
  - Distribution spans the first four AQI categories, though values still skew heavily toward “Good.” RP is the only site capturing a meaningful number of hours in the “USG” and “Unhealthy” ranges.
  - Among the three FEM instruments (1020, 1022, Thermo), measurements are largely aligned but diverge slightly during higher concentration episodes. The FEM average smooths out hourly variability and serves as the gold standard for evaluation and modeling for this project.

- **NC Site:**
  - PM₂.₅ is tightly clustered in the “Good” AQI range for all monitors, with little spread and FEM values rarely exceeding ~10 µg/m³.
  - A brief cold snap in Q1 2025 (with temperatures reaching -20 to -30°F) caused outages and degraded performance in both FEM and Modulair data. These anomalies contributed to unusually poor agreement between the sensor and FEM, limiting this site's utility for analysis and modeling.

- **LT Site:**
  - Like NC, air quality conditions at LT were clean throughout the sampling period. FEM values rarely surpassed 10 µg/m³, with modest excursions into “Moderate.”
  - All sensors—including the PurpleAir—tracked reasonably well at low concentrations, though signal variation increased slightly for the Modulair.

## 2. Accuracy of Modulair vs. FEM by Site

To assess the raw Modulair PM₂.₅ accuracy, we calculated a suite of sensor evaluation metrics including R², RMSE, MAE, normalized RMSE (NRMSE), and AQI match frequency using hourly data. These metrics are benchmarked against both:

- **EPA (2021) NSIM performance targets** for PM₂.₅ field sensors.
- **QuantAQ’s own Modulair fact sheet**, which reports average hourly performance metrics.

#### Overall Accuracy Metrics (Hourly)

| Site | R²    | MAE (µg/m³) | RMSE (µg/m³) | NRMSE (%) | AQI Match (%) | EPA Target (Hourly)       | QuantAQ Target (Hourly)  |
|------|-------|--------------|--------------|------------|----------------|----------------------------|---------------------------|
| RP   | 0.67  | 3.83         | 6.36         | 77.0       | 82.1%          | R² > 0.70, RMSE < 7        | R² = 0.92, RMSE = 3.2     |
| NC   | 0.017 | 3.78         | 9.00         | 475.7      | 79.8%          | —                          | —                         |
| LT   | 0.029 | 2.97         | 4.37         | 152.4      | 85.3%          | —                          | —                         |

- At **RP**, raw Modulair values come close to meeting EPA deployment thresholds, particularly RMSE (<7). R² is slightly low (0.67), and performance falls short of QuantAQ’s claimed R² = 0.92 and RMSE = 3.2.
- At **NC**, accuracy breaks down entirely. The extremely low correlation and high NRMSE (~476%) point to data quality concerns during and after equipment failure.
- At **LT**, RMSE is within EPA guidelines but again falls short of manufacturer claims. High NRMSE (~152%) stems from the low overall concentration range.

These results justify prioritizing RP for modeling use. It is the only site with data quality and variability sufficient to support sensor calibration.

#### Direction of Bias by AQI Category

We also analyzed how sensor error varied by AQI level. Below is a summary of the average bias (FEM − Modulair), by AQI category and site:

| Site | AQI Category | Average Error (µg/m³) | Bias Direction         |
|------|---------------|------------------------|------------------------|
| RP   | Good          | +0.15                  | Slight Underestimate   |
| RP   | Moderate      | -0.76                  | Overestimate           |
| RP   | USG           | -8.47                  | Strong Overestimate    |
| RP   | Unhealthy     | -35.47                 | Strong Overestimate    |
| NC   | Good          | -2.30                  | Overestimate           |
| LT   | Good          | +1.98                  | Underestimate          |
| LT   | Moderate      | +1.47                  | Underestimate          |

Key takeaways:

- At **RP**, error is near-zero in clean conditions but increases sharply with pollution. Modulair overestimates significantly in the “Unhealthy” category.
- At **LT**, the sensor underestimates across the board, but the overall bias is small in magnitude.
- At **NC**, consistent overestimation occurs even in clean air—again suggesting data reliability issues.

These stratified findings confirm the **nonlinear nature of bias**, particularly at higher concentrations. They also show the importance of using AQI-stratified metrics in addition to overall accuracy.

## 3. Modeling Implications

- **Site Selection:**
  - RP is the most viable site for model development due to its broader PM₂.₅ range, sufficient AQI representation, and data integrity.
  - NC and LT are not appropriate for model fitting but may serve in future model validation under low-concentration scenarios.

- **Sensor Performance Summary:**
  - Modulair sensor readings are reasonably aligned with FEM values at RP, especially under 35 µg/m³. However, they substantially overestimate in higher pollution events.
  - PurpleAir values (which are already corrected using the EPA US-wide formula) serve as a useful benchmark but do not consistently outperform raw Modulair data—particularly at RP.

- **Modeling Approach:**
  - Our modeling strategy focuses on improving raw Modulair performance through bias correction. A base linear model is a reasonable start.
  - Given known scattering artifacts due to RH, its inclusion as an additive or interaction term is justified.
  - Future exploration may include polynomial or piecewise terms to address underperformance in high AQI categories, though complexity must be weighed against operational deployment constraints.

These conclusions justify our focus on the RP site and provide the foundation for the modeling script that follows. They also highlight key challenges and opportunities for improving sensor performance through light-touch statistical correction.

```{r}
# Save All Updated kable() Tables to Excel Files for report 

library(writexl)

# Define data folder
data_folder <- here("data")

# Save each updated table individually

# 1. Summary of Observations and AQI Category Counts
write_xlsx(combined_summaries, file.path(data_folder, "summary_aqi_counts.xlsx"))

# 2. Descriptive Statistics for RP, NC, LT
write_xlsx(rp_stats, file.path(data_folder, "rp_descriptive_stats.xlsx"))
write_xlsx(nc_stats, file.path(data_folder, "nc_descriptive_stats.xlsx"))
write_xlsx(lt_stats, file.path(data_folder, "lt_descriptive_stats.xlsx"))

# 3. Overall Accuracy Metrics (Combined)
write_xlsx(combined_overall_table, file.path(data_folder, "combined_overall_accuracy_metrics.xlsx"))

# 4. Stratified Accuracy Metrics (Combined)
write_xlsx(combined_stratified_table, file.path(data_folder, "combined_stratified_accuracy_metrics.xlsx"))

# 5. Overall Accuracy Metrics by Site
write_xlsx(rp_metrics_table, file.path(data_folder, "rp_overall_accuracy_metrics.xlsx"))
write_xlsx(nc_metrics_table, file.path(data_folder, "nc_overall_accuracy_metrics.xlsx"))
write_xlsx(lt_metrics_table, file.path(data_folder, "lt_overall_accuracy_metrics.xlsx"))

# 6. Stratified Accuracy Metrics by Site
write_xlsx(rp_stratified_table, file.path(data_folder, "rp_stratified_accuracy_metrics.xlsx"))
write_xlsx(nc_stratified_table, file.path(data_folder, "nc_stratified_accuracy_metrics.xlsx"))
write_xlsx(lt_stratified_table, file.path(data_folder, "lt_stratified_accuracy_metrics.xlsx"))

# Confirmation Message
cat("✅ All updated tables printed with kable() have been saved successfully to the data folder.\n")

```


