---
title: "Exploratory Analysis of MT DEQ FEM/Modulair PM2.5 Collocation Data"
author: "Mason Dow"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

This document performs an exploratory analysis of collocated FEM BAM and Modulair PM2.5 datasets from three MT DEQ sites (RP, NC, and LT). The objectives are to:

- Generate summary tables of descriptive statistics (min, max, mean, standard deviation, etc.) for each parameter for each site.
- Visualize PM2.5 trends and distributions by site
- Compute AQI categories for PM2.5 using the AirMontitor package.
- Calculate a comprehensive set of accuracy metrics (including R², MAE, RMSE, normalized RMSE, regression slope, and intercept), overall and by AQI category for   each site.

# Setup

```{r setup, message=FALSE, warning=FALSE, echo=FALSE}

# Load necessary libraries
library(tidyverse)
library(readxl)
library(lubridate)
library(scales)
library(knitr)
library(AirMonitor)  
library(here)
library(GGally)

source(here("functions", "analysis_functions.R"))

theme_set(theme_minimal())
```


# Load Data

```{r}
# Load data from data folder 
lt_data <- read_excel("../data/LT_data.xlsx")
nc_data <- read_excel("../data/NC_data.xlsx")
rp_data <- read_excel("../data/RP_data.xlsx")
```

# Data Transformations

## Calculate 24 hr. Averages

```{r}
# uses averaging_threshold and daily_averages functions from function file
# applies 75% (18 hr.) threshold to hourly values to caclulate daily averages 

# Apply the function to each dataset:
rp_daily <- daily_averages(rp_data)
nc_daily <- daily_averages(nc_data)
lt_daily <- daily_averages(lt_data)

head(rp_daily)
```


## Color Code and Column Name Mapping by Monitor

```{r}
# Color scheme for graphing by monitor types

monitor_colors <- c("1020" = "#56B4E9", 
                    "1022" = "#D55E00", 
                    "Thermo" = "#009E73", 
                    "FEM Average" = "red",
                    "Modulair" = "#E69F00", 
                    "PurpleAir" = "purple")

# Define recoding maps for each site to convert raw column names to monitor labels

rp_recode <- c("rp_1020" = "1020",
               "rp_1022" = "1022",
               "rp_thermo" = "Thermo",
               "rp_fem_avg" = "FEM Average",
               "rp_modulair_pm25" = "Modulair",
               "rp_purpleair_cf1" = "PurpleAir")

nc_recode <- c("nc_1020" = "1020",
               "nc_modulair_pm25" = "Modulair",
               "nc_purpleair_cf1" = "PurpleAir")

lt_recode <- c("lt_thermo" = "Thermo",
               "lt_modulair_pm25" = "Modulair",
               "lt_purpleair_cf1" = "PurpleAir")
```

## Define AQI Breakpoints 

```{r}
# Standard US daily PM2.5 AQI breakpoints (24-hr, updated 2024)
# Good: 0.0 - 9.0
# Moderate: 9.1 - 35.4
# Unhealthy for Sensitive Groups: 35.5 - 55.4 
# Unhealthy: 55.5 - 125.4
# Very Unhealthy: 125.5 - 225.4
# Hazardous: 225.5 +

aqi_breaks  <- c(9, 35.4, 55.4, 125.4, 225.4)
aqi_colors  <- c("green", "yellow", "orange", "red", "purple")
```

## Calculate Corresponding AQI Categories by site and PM2.5 parameter 

```{r}
# Hourly Data

# Add AQI categories using AirMonitor's aqiCategories() function.
rp_data <- rp_data %>%
  mutate(rp_modulair_aqi = aqiCategories(rp_modulair_pm25, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         rp_purpleair_aqi = aqiCategories(rp_purpleair_cf1, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         rp_fem_aqi = aqiCategories(rp_fem_avg, pollutant = "PM2.5", NAAQS = "PM2.5_2024"))

nc_data <- nc_data %>%
  mutate(nc_modulair_aqi = aqiCategories(nc_modulair_pm25, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         nc_purpleair_aqi = aqiCategories(nc_purpleair_cf1, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         nc_fem_aqi = aqiCategories(nc_1020, pollutant = "PM2.5", NAAQS = "PM2.5_2024"))

lt_data <- lt_data %>%
  mutate(lt_modulair_aqi = aqiCategories(lt_modulair_pm25, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         lt_purpleair_aqi = aqiCategories(lt_purpleair_cf1, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         lt_fem_aqi = aqiCategories(lt_thermo, pollutant = "PM2.5", NAAQS = "PM2.5_2024"))


```

```{r}
# Daily Data

# Add AQI categories using AirMonitor's aqiCategories() function.
rp_daily <- rp_daily %>%
  mutate(rp_modulair_aqi = aqiCategories(rp_modulair_pm25, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         rp_purpleair_aqi = aqiCategories(rp_purpleair_cf1, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         rp_fem_aqi = aqiCategories(rp_fem_avg, pollutant = "PM2.5", NAAQS = "PM2.5_2024"))

nc_daily <- nc_daily %>%
  mutate(nc_modulair_aqi = aqiCategories(nc_modulair_pm25, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         nc_purpleair_aqi = aqiCategories(nc_purpleair_cf1, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         nc_fem_aqi = aqiCategories(nc_1020, pollutant = "PM2.5", NAAQS = "PM2.5_2024"))

lt_daily <- lt_daily %>%
  mutate(lt_modulair_aqi = aqiCategories(lt_modulair_pm25, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         lt_purpleair_aqi = aqiCategories(lt_purpleair_cf1, pollutant = "PM2.5", NAAQS = "PM2.5_2024"),
         lt_fem_aqi = aqiCategories(lt_thermo, pollutant = "PM2.5", NAAQS = "PM2.5_2024"))


```

## Create Combined Dataset with Consitent Column Names

```{r}
# For all datasets, rename site-specific columns to common names via *_combined tables

rp_combined <- rp_data %>%
  rename(
    fem_avg = rp_fem_avg,            # FEM measurement
    mod_pm25 = rp_modulair_pm25,       # Modulair PM2.5 measurement
    mod_rh = rp_modulair_rh,           # Modulair measured RH
    mod_tempc = rp_modulair_tempc,     # Modulair measured ambient temperature (°C)
    fem_aqi = rp_fem_aqi,              # FEM AQI category
    mod_aqi = rp_modulair_aqi          # Modulair AQI category
  )

nc_combined <- nc_data %>%
  rename(
    fem_avg = nc_1020,          # FEM measurement from NC
    mod_pm25 = nc_modulair_pm25,  # Modulair PM2.5 measurement
    fem_aqi = nc_fem_aqi,         # FEM AQI category
    mod_aqi = nc_modulair_aqi,     # Modulair AQI category
    mod_rh = nc_modulair_rh,       # Modulair measured RH
    mod_tempc = nc_modulair_tempc,     # Modulair measured ambient temperature (°C)
  )

lt_combined <- lt_data %>%
  rename(
    fem_avg = lt_thermo,         # FEM measurement from LT
    mod_pm25 = lt_modulair_pm25,   # Modulair PM2.5 measurement
    fem_aqi = lt_fem_aqi,          # FEM AQI category
    mod_aqi = lt_modulair_aqi,      # Modulair AQI category
    mod_rh = lt_modulair_rh,        # Modulair measured RH
    mod_tempc = lt_modulair_tempc,     # Modulair measured ambient temperature (°C)
  )

# Combine the datasets into one dataset
combined_data <- bind_rows(rp_combined, nc_combined, lt_combined)

# Select only the columns that follow our standardized naming scheme (plus datetime and siteid)
combined_data <- combined_data %>%
  select(siteid, datetime, fem_avg, mod_pm25, mod_rh, mod_tempc, fem_aqi, mod_aqi)


# Check the first few rows to verify the merge worked as expected.
head(combined_data)

```


# Calculate Descriptive Statistics

```{r}
# Summarize observations by site, AQI Category 

# uses summarize_counts function from function file 
# adjust fem_col, mod_col, aqi_col as appropriate for each dataset.

# For RP site:
rp_summary <- summarize_counts(
  data     = rp_data,
  fem_col  = "rp_fem_avg",       # or rp_fem_weighted_avg if that's your main reference
  mod_col  = "rp_modulair_pm25",
  aqi_col  = "rp_fem_aqi"
)

# For NC site 
nc_summary <- summarize_counts(
  data     = nc_data,
  fem_col  = "nc_1020",
  mod_col  = "nc_modulair_pm25",
  aqi_col  = "nc_fem_aqi"
)

# For LT site:
lt_summary <- summarize_counts(
  data     = lt_data,
  fem_col  = "lt_thermo",
  mod_col  = "lt_modulair_pm25",
  aqi_col  = "lt_fem_aqi"
)

# Combine site-level summaries into one table, print w/ kable
combined_summaries <- bind_rows(rp_summary, nc_summary, lt_summary)

kable(
  combined_summaries,
  caption = "Summary of Observations and AQI Category Counts/Percentages by Site"
)

```


```{r}
# build out descriptive statistics tables by site
# uses descriptive_stats function from function file 

rp_stats <- descriptive_stats(rp_data)
nc_stats <- descriptive_stats(nc_data)
lt_stats <- descriptive_stats(lt_data)

kable(rp_stats, caption = "Descriptive Statistics - RP Site")
kable(nc_stats, caption = "Descriptive Statistics - NC Site")
kable(lt_stats, caption = "Descriptive Statistics - LT Site")
```

# Data Visualizations and Explorations 

## Boxplots of PM2.5 Concentration Distributions

```{r}
# create boxplots of PM2.5 distributions by site and monitor
# uses stack_monitor_data function from function file

rp_stacked <- stack_monitor_data(rp_data, "datetime", names(rp_recode), rp_recode) # create "stacked" monitor data to invert columns and rows for plotting
ggplot(rp_stacked, aes(x = monitor, y = pm25, fill = monitor)) +
  geom_boxplot() +
  scale_fill_manual(values = monitor_colors) +
  labs(title = "PM2.5 Distribution at RP Site", x = "Monitor", y = "PM2.5 (µg/m³)") +
  geom_hline( # Add annotated line at each AQI breakpoint
    yintercept = aqi_breaks,
    color = aqi_colors,
    linetype = "dashed"
  ) +
  annotate("text", x = Inf, y = 12,
           label = "Good",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 35.4,
           label = "Moderate",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 55.4,
           label = "USG",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 150.4,
           label = "Unhealthy",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 250.4,
           label = "V. Unhealthy",
           vjust = -0.5, hjust = 1.1) +
  theme_minimal() +
  theme(legend.position = "none")

nc_stacked <- stack_monitor_data(nc_data, "datetime", names(nc_recode), nc_recode)
ggplot(nc_stacked, aes(x = monitor, y = pm25, fill = monitor)) +
  geom_boxplot() +
  scale_fill_manual(values = monitor_colors) +
  labs(title = "PM2.5 Distribution at NC Site", x = "Monitor", y = "PM2.5 (µg/m³)") +
  geom_hline( 
    yintercept = aqi_breaks,
    color = aqi_colors,
    linetype = "dashed"
  ) +
  annotate("text", x = Inf, y = 12,
           label = "Good",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 35.4,
           label = "Moderate",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 55.4,
           label = "USG",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 150.4,
           label = "Unhealthy",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 250.4,
           label = "V. Unhealthy",
           vjust = -0.5, hjust = 1.1) +
  theme_minimal() +
  theme(legend.position = "none")

lt_stacked <- stack_monitor_data(lt_data, "datetime", names(lt_recode), lt_recode)
ggplot(lt_stacked, aes(x = monitor, y = pm25, fill = monitor)) +
  geom_boxplot() +
  scale_fill_manual(values = monitor_colors) +
  labs(title = "PM2.5 Distribution at LT Site", x = "Monitor", y = "PM2.5 (µg/m³)") +
 geom_hline( 
    yintercept = aqi_breaks,
    color = aqi_colors,
    linetype = "dashed"
  ) +
  annotate("text", x = Inf, y = 12,
           label = "Good",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 35.4,
           label = "Moderate",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 55.4,
           label = "USG",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 150.4,
           label = "Unhealthy",
           vjust = -0.5, hjust = 1.1) +
  annotate("text", x = Inf, y = 250.4,
           label = "V. Unhealthy",
           vjust = -0.5, hjust = 1.1) +
  theme_minimal() +
  theme(legend.position = "none")

```

## Time Series Line Graphs of 24 hour averaged PM2.5 Concentrations by Site

```{r}

final_date <- as.Date("2025-02-27")

# RP Site

ggplot(rp_daily, aes(x = date)) +
  geom_line(aes(y = rp_1020, color = "1020")) +
  geom_line(aes(y = rp_1022, color = "1022")) +
  geom_line(aes(y = rp_thermo, color = "Thermo")) +
  geom_line(aes(y = rp_modulair_pm25, color = "Modulair")) +
  geom_line(aes(y = rp_purpleair_cf1, color = "PurpleAir")) +
  # add AQI lines
  geom_hline(
    yintercept = aqi_breaks,
    color = aqi_colors,
    linetype = "dashed"
  ) +
  annotate("text", x = final_date + days(50), y = 12,     label = "Good",         vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(50), y = 35.4,   label = "Moderate",     vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(50), y = 55.4,   label = "USG",          vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(50), y = 150.4,  label = "Unhealthy",    vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(50), y = 250.4,  label = "V. Unhealthy", vjust = -0.5, hjust = 1.1) +
  scale_color_manual(values = monitor_colors) +
  labs(
    title = "PM2.5 Time Series at RP Site",
    x = "Date",
    y = "PM2.5 (µg/m³)",
    color = "Monitor Type"
  ) +
  theme_minimal()

# NC Site
ggplot(nc_daily, aes(x = date)) +
  geom_line(aes(y = nc_1020, color = "1020")) +
  geom_line(aes(y = nc_modulair_pm25, color = "Modulair")) +
  geom_line(aes(y = nc_purpleair_cf1, color = "PurpleAir")) +
  geom_hline(
    yintercept = aqi_breaks,
    color = aqi_colors,
    linetype = "dashed"
  ) +
  annotate("text", x = final_date + days(10), y = 12,     label = "Good",         vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(10), y = 35.4,   label = "Moderate",     vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(10), y = 55.4,   label = "USG",          vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(10), y = 150.4,  label = "Unhealthy",    vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(10), y = 250.4,  label = "V. Unhealthy", vjust = -0.5, hjust = 1.1) +
  scale_color_manual(values = monitor_colors) +
  labs(
    title = "PM2.5 Time Series at RP Site",
    x = "Date",
    y = "PM2.5 (µg/m³)",
    color = "Monitor Type"
  ) +
  theme_minimal()

# LT Site
ggplot(lt_daily, aes(x = date)) +
  geom_line(aes(y = lt_thermo, color = "Thermo")) +
  geom_line(aes(y = lt_modulair_pm25, color = "Modulair")) +
  geom_line(aes(y = lt_purpleair_cf1, color = "PurpleAir")) +
  geom_hline(
    yintercept = aqi_breaks,
    color = aqi_colors,
    linetype = "dashed"
  ) +
  annotate("text", x = final_date + days(10), y = 12,     label = "Good",         vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(10), y = 35.4,   label = "Moderate",     vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(10), y = 55.4,   label = "USG",          vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(10), y = 150.4,  label = "Unhealthy",    vjust = -0.5, hjust = 1.1) +
  annotate("text", x = final_date + days(10), y = 250.4,  label = "V. Unhealthy", vjust = -0.5, hjust = 1.1) +
  scale_color_manual(values = monitor_colors) +
  labs(
    title = "PM2.5 Time Series at RP Site",
    x = "Date",
    y = "PM2.5 (µg/m³)",
    color = "Monitor Type"
  ) +
  theme_minimal()

```

## Scatterplot w/ Line of Best Fit & 1:1 by Site

```{r}

plot_rp <- plot_fit(rp_data, mod_col = "rp_modulair_pm25", fem_col = "rp_fem_avg", site_label = "RP")
print(plot_rp)

plot_nc <- plot_fit(nc_data, mod_col = "nc_modulair_pm25", fem_col = "nc_1020", site_label = "NC")
print(plot_nc)

plot_lt <- plot_fit(lt_data, mod_col = "lt_modulair_pm25", fem_col = "lt_thermo", site_label = "LT")
print(plot_lt)

```
## Correlation Matrix by Modeling Variable 

```{r}
# Create correlation matrix with all variables to be potentially included in modeling using combined dataset

# - fem_avg: FEM PM2.5 measurement (true value)
# - mod_pm25: Modulair PM2.5 measurement (raw)
# - mod_rh: Modulair relative humidity
# - mod_tempc: Modulair ambient temperature (°C)

model_vars <- combined_data %>%
  select(fem_avg, mod_pm25, mod_rh, mod_tempc)

# Calculate and visualize the correlation matrix using ggpairs
ggpairs(model_vars,
        lower = list(continuous = wrap("smooth", alpha = 0.3, size = 0.1)),
        upper = list(continuous = wrap("cor", size = 4)),
        diag = list(continuous = "barDiag")) +
  labs(title = "Correlation Matrix of Candidate Modeling Variables")
```

### Matrix Interpretation:

- Moderate correlation between reference and raw sensor PM2.5 (r ∼ 0.77) suggests the sensor is “in the ballpark” but needs a bias correction.

- Minimal linear correlation between PM2.5 and RH/temperature does not rule out the possibility that adding an RH or temperature term can significantly improve your final model—particularly if the sensor’s error depends on meteorology.

- Strong collinearity between temperature and humidity (r ~ −0.66) means you should be mindful about including both in a single linear regression. In some cases, one meteorological term (often RH) captures most of the sensor’s bias.
 
# Measures of Sensor Accuracy

## Metrics by Site & AQI Category
 
```{r}
# For Loop to calculate accuracy metrics functions for each site & overall 

datasets <- list(rp = rp_data, nc = nc_data, lt = lt_data)
results <- list()

col_info <- list(
  rp = list(fem = "rp_fem_avg", mod = "rp_modulair_pm25", fem_aqi = "rp_fem_aqi", mod_aqi = "rp_modulair_aqi"),
  nc = list(fem = "nc_1020", mod = "nc_modulair_pm25", fem_aqi = "nc_fem_aqi", mod_aqi = "nc_modulair_aqi"),
  lt = list(fem = "lt_thermo", mod = "lt_modulair_pm25", fem_aqi = "lt_fem_aqi", mod_aqi = "lt_modulair_aqi")
)

for (site in names(datasets)) {
  results[[site]] <- calculate_all_metrics(
    data = datasets[[site]],
    fem_col = col_info[[site]]$fem,
    mod_col = col_info[[site]]$mod,
    fem_aqi_col = col_info[[site]]$fem_aqi,
    mod_aqi_col = col_info[[site]]$mod_aqi
  )
}

# bind tables, print results 

overall_table <- bind_rows(lapply(names(results), function(site) {
  results[[site]]$overall %>% mutate(site = site)
}))
overall_table <- overall_table %>% select(site, everything())
kable(overall_table, caption = "Overall Accuracy Metrics by Site (FEM vs. Modulair)")

for (site in names(results)) {
  print(kable(results[[site]]$stratified, caption = paste("Accuracy Metrics by AQI Category -", site, "Site")))
}
```
 
## Calculate Overall Cccuracy Metrics 

```{r}
# apply accuracy metrics functions to combined dataset

combined_metrics <- calculate_all_metrics(
  data = combined_data,
  fem_col = "fem_avg",
  mod_col = "mod_pm25",
  fem_aqi_col = "fem_aqi",
  mod_aqi_col = "mod_aqi"
)

# Display the overall metrics for the combined dataset
print(combined_metrics$overall)

# Display the stratified accuracy metrics by AQI category for the combined dataset
kable(combined_metrics$stratified, caption = "Accuracy Metrics by AQI Category - Combined Data")
```

