---
title: "Secondary Data Cleaning and Preprocessing"
author: "Mason"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Required Libraries
```{r}
library(tidyverse)
library(lubridate)
library(readxl)
```

## Load Data
```{r}
# Define file paths (update these paths as needed)
lt_data <- read_excel("../data/LT_data.xlsx")
nc_data <- read_excel("../data/NC_data.xlsx")
rp_data <- read_excel("../data/RP_data.xlsx")

# Standardizing column names to lowercase
lt_data <- rename_with(lt_data, tolower)
nc_data <- rename_with(nc_data, tolower)
rp_data <- rename_with(rp_data, tolower)
```

## Convert DateTime and Check Structure
```{r}
lt_data <- lt_data %>% mutate(datetime = as_datetime(datetime))
nc_data <- nc_data %>% mutate(datetime = as_datetime(datetime))
rp_data <- rp_data %>% mutate(datetime = as_datetime(datetime))

# View data structures
glimpse(lt_data)
glimpse(nc_data)
glimpse(rp_data)
```

# Step 1: Data Screening and Initial Cleaning

## 1.1 Check Time Gaps
```{r}
check_time_gaps <- function(data, site_name) {
  data %>%
    arrange(datetime) %>%  # Ensures data is sorted chronologically
    mutate(time_diff = difftime(datetime, lag(datetime), units = "hours")) %>%  # Computes time difference
    count(time_diff, name = "count") %>%  # Counts occurrences of each unique time gap
    mutate(site = site_name)  # Adds a site identifier
}


lt_gaps <- check_time_gaps(lt_data, "LT")
nc_gaps <- check_time_gaps(nc_data, "NC")
rp_gaps <- check_time_gaps(rp_data, "RP")

time_gaps <- bind_rows(lt_gaps, nc_gaps, rp_gaps)
time_gaps
```

## 1.2 Missing Data Analysis
```{r}
compute_missing_summary <- function(data, site_name) {
  fem_col <- names(data)[str_detect(names(data), "thermo|1020|1022|fem_avg")] %>% first() # Selects the first matching FEM column
  mod_col <- names(data)[str_detect(names(data), "modulair_pm25")] %>% first()
  rh_col <- names(data)[str_detect(names(data), "modulair_rh")] %>% first()
  temp_col <- names(data)[str_detect(names(data), "modulair_tempc")] %>% first()
  
  data %>%
    summarise(
      site = site_name,
      missing_pm25_fem = sum(is.na(.data[[fem_col]])),
      missing_pm25_mod = sum(is.na(.data[[mod_col]])),
      missing_rh = sum(is.na(.data[[rh_col]])),
      missing_temp = sum(is.na(.data[[temp_col]]))
    )
}

lt_missing <- compute_missing_summary(lt_data, "LT")
nc_missing <- compute_missing_summary(nc_data, "NC")
rp_missing <- compute_missing_summary(rp_data, "RP")

missing_table <- bind_rows(lt_missing, nc_missing, rp_missing)
missing_table
```

## 1.3 Outlier Detection
```{r}
# Function to visualize outliers with boxplots
plot_outliers <- function(data, column, site_name) {
  ggplot(data, aes(y = .data[[column]])) +
    geom_boxplot() +
    labs(title = paste("Outlier Detection for", column, "at", site_name),
         y = column) +
    theme_minimal()
}

# Generate outlier plots for PM2.5 concentrations
plot_outliers(lt_data, "lt_modulair_pm25", "LT")
plot_outliers(nc_data, "nc_modulair_pm25", "NC")
plot_outliers(rp_data, "rp_modulair_pm25", "RP")

# Generate histograms to visualize distributions
plot_histogram <- function(data, column, site_name) {
  ggplot(data, aes(x = .data[[column]])) +
    geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
    labs(title = paste("Histogram of", column, "at", site_name),
         x = column, y = "Frequency") +
    theme_minimal()
}

plot_histogram(lt_data, "lt_modulair_pm25", "LT")
plot_histogram(nc_data, "nc_modulair_pm25", "NC")
plot_histogram(rp_data, "rp_modulair_pm25", "RP")
```

## Conclusion
This section performs initial **data screening** by:
- Checking for **time gaps** in hourly data.
- Summarizing **missing data** across key variables.
- Identifying **outliers** using boxplots and histograms.


